---
title: "final"
author: "Joshua Zhang"
date: "5-26-2024"
---

```{r}
library(tidyverse)
```

## Target Questions
- What are the impacts of infecting various diseases like malaria, pneumonia, and diarrheal diseases in contributing to child mortality? Which factor plays the biggest role in triggering mortality?
- How does the country’s economic conditions and income levels affect child mortality?
- How does age impact child mortality, especially when it comes to diseases?
- What is the current global child mortality rate? How is the mortality rate distributed across different continents?



## Viewing Datasets

#### 1. Loading Datasets
```{r}
cause_of_death <- read_delim("causes-of-death-in-children.csv")
by_age <- read_delim("child-deaths-by-age.csv")
mortality_fertility_income <- read_delim("MortalityFertilityIncome.csv")

head(cause_of_death)
head(by_age)
head(mortality_fertility_income)
```

#### 2. Viewing Data Columns
```{r}
cat("Cause of Death\n")
print(names(cause_of_death))
cat("\n")
cat("Death By Age\n")
print(names(by_age))
cat("Mortality Fertility And Income\n")
print(names(mortality_fertility_income))
```
Year looks like a similar column we could use in all the datasets. We will need to pivot the mortality fertility income (MFI) dataset to be longer.


#### 3. Splitting the Mortality Fertility Income (MFI) dataframe
```{r}
mortality <- mortality_fertility_income %>%
  select(starts_with("m"))

fertility <- mortality_fertility_income %>%
  select(starts_with("f"))

income <- mortality_fertility_income %>%
  select(starts_with("i"))

names(mortality) <- substring(names(mortality), 2)
names(fertility) <- substring(names(fertility), 2)
names(income) <- substring(names(income), 2)

mortality <- cbind(mortality, mortality_fertility_income[,c('Country Name', 'Country Code', 'Region')])
fertility <- cbind(fertility, mortality_fertility_income[,c('Country Name', 'Country Code', 'Region')])
income <- cbind(income, mortality_fertility_income[,c('Country Name', 'Country Code', 'Region')])
```

#### 4. Making the income dataframe long instead of wide
```{r}
mortality_longer <- pivot_longer(mortality, -c("Country Code", "Country Name", "Region"), names_to="Year", values_to="Mortality")

income_longer <- pivot_longer(income, -c("Country Code", "Country Name", "Region"), names_to="Year", values_to="Income")
```



#### 5. Mortality Vs Income

```{r}
m_i <- merge(mortality_longer, income_longer)

m_i$Year <- as.integer(m_i$Year)

m_i <- m_i %>%
  filter(!is.na(Mortality)) %>%
  filter(!is.na(Income)) %>%
  filter(`Country Code` == "USA")

ggplot(m_i, aes(x = Mortality, y = Income, color = factor(floor(Year / 10) * 10))) +
  geom_point() 
```


## Child Mortality Causes

#### 1. Changing dataset names.
```{r}
income_renamed <- rename(income_longer, Code="Country Code")
print(names(income_renamed))
```

#### 2. Merging the two datasets, reshape longer
```{r}
income_renamed$Year <- as.integer(income_renamed$Year)
disease_income <- merge(income_renamed, cause_of_death)
disease_income_longer <- pivot_longer(disease_income, -c("Code", "Country Name", "Region", "Income", "Entity", "Year"), names_to="Disease", values_to="Deaths")
```

#### 3. Plot Disease vs Income
```{r}
disease_filtered <- filter(disease_income_longer[disease_income_longer$Code == "AFG", ])

ggplot(disease_filtered, aes(x = Income, y = Deaths, color = Disease)) +
  geom_point() +
  theme(legend.position="none")
```


#### 4. Loading population data
```{r}
populations <- read_delim("population.csv")
populations[, c("Indicator Name", "Indicator Code")] <- list(NULL)
head(populations)
```

#### 5. Pivot population longer
```{r}
populations_longer <- pivot_longer(populations, -c("Country Code", "Country Name"), names_to="Year", values_to="Population")

populations_longer <- rename(populations_longer, Code="Country Code")
```


#### 6. Disease Impact WorldWide
```{r}
total_disease <- disease_filtered %>% 
  group_by(Disease, Year) %>%
  summarise(total_deaths = sum(Deaths, na.rm=TRUE))


ggplot(total_disease, aes(x = Year, y = total_deaths, color = Disease)) +
  geom_point() +
  theme(legend.position="none")
```